---
title: "Code-Through: Data Wrangling with dplyr and the Star Wars Dataset"
author: "Gregorio Salcedo"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary packages
library(dplyr)
library(ggplot2)
library(gganimate)
```

## Introduction
This tutorial will demonstrate how to use `dplyr` for data wrangling using the `starwars` dataset from the `dplyr` package.

```{r load-data}
# Load the Star Wars dataset
starwars <- dplyr::starwars
# View the first few rows
head(starwars)
```

## Selecting Columns
We can use `select()` to choose specific columns from the dataset.

```{r select-columns}
starwars_selected <- starwars %>% 
  select(name, height, mass, species)

head(starwars_selected)
```

## Filtering Rows
Use `filter()` to extract characters meeting specific conditions.

```{r filter-data}
# Characters taller than 200cm
starwars_tall <- starwars %>% 
  filter(height > 200)

starwars_tall
```

## Arranging Data
We can sort data with `arrange()`.

```{r arrange-data}
# Arrange characters by mass in descending order
starwars_sorted <- starwars %>% 
  arrange(desc(mass))

head(starwars_sorted)
```

## Mutating (Adding New Columns)
Create new columns using `mutate()`.

```{r mutate-data}
# Add a BMI column
starwars <- starwars %>% 
  mutate(BMI = mass / ((height / 100) ^ 2))

select(starwars, name, mass, height, BMI)
```

## Summarizing Data
Use `summarize()` to compute summary statistics.

```{r summarize-data}
# Average height of characters
starwars_summary <- starwars %>% 
  summarise(average_height = mean(height, na.rm = TRUE))

starwars_summary
```

## Grouping Data
We can analyze data by groups using `group_by()`.

```{r group-by}
# Average height by species
starwars_grouped <- starwars %>% 
  group_by(species) %>% 
  summarise(avg_height = mean(height, na.rm = TRUE)) %>%
  arrange(desc(avg_height))

head(starwars_grouped)
```

## Visualization: Plot of Height vs. Mass
Let's create a scatter plot showing the relationship between height and mass.

```{r plot-height-mass}
# Scatter plot of height vs. mass
plot <- ggplot(starwars, aes(x = height, y = mass, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "Height vs. Mass of Star Wars Characters", x = "Height (cm)", y = "Mass (kg)") +
  theme_minimal()

plot
```

## Animated Plot: Character Growth
We'll create an animated plot to show how characters are distributed by height over time.

```{r animate-height, warning=FALSE}
# Animation showing height distribution by species
animated_plot <- ggplot(starwars, aes(x = height, fill = species)) +
  geom_histogram(binwidth = 10, alpha = 0.7, position = "identity") +
  labs(title = "Distribution of Star Wars Character Heights", x = "Height (cm)", y = "Count") +
  theme_minimal() +
  transition_states(species, transition_length = 2, state_length = 1)

animated_plot
```


## Conclusion
This tutorial demonstrated how to use `dplyr` to manipulate data efficiently. The `starwars` dataset is great for practicing filtering, selecting, and summarizing data. Happy coding!
